Effectful Semantics for Substructural SSA

### 2023-11-17 pldi24/paper.tex
Tap

One issue with the above definition is that, just like for regular premonoidal
categories, it is difficult to distinguish between those morphisms that are
"truly" central versus those which are just central "by accident", as we only
want to require any notion of functor we develop subsequently to preserve the
former. We can solve this the same way, by simply extending the notion of
effectful category appropriately as follows:
\begin{definition}[Susbtructural Effectful Category]
  A \textbf{substructural effectful category} \(\mc{C}\) consists of
  \begin{itemize}
    \item A substructural monoidal category \(\mc{C}_1\); we call the morphisms
    in this category \textbf{central} or \textbf{linear}
    \item A substructural premonoidal category \(\mc{C}_0\) sharing the same
    affine and relevant objects as \(\mc{C}_1\)
    \item An identity-on-objects functor \(\upg{\cdot}{}: \mc{C}_1^\varnothing
    \to \mc{C}_0^\varnothing\) preserving all substructural premonoidal
    structure, i.e., preserves all symmetries, associators, unitors, drops, and
    splits and sends \(\mc{C}_1^q\) to \(\mc{C}_0^q\). In particular,
    restricting \(\upg{\cdot}{}\) appropriately makes all \(\mc{C}_1^q \to
    \mc{C}_0^q\) into effectful categories.
  \end{itemize}
  We define \(\upg{f}{0}\) to be \(\upg{f}{}\) for morphisms in \(\mc{C}_1^q\)
  and simply \(f\) for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) to simply
  be the identity functor on \(\mc{C}_1^q\).

  Morphisms in \(\mc{C}_p^{\ms{r}}\) are called \textbf{copyable}, morphisms
  in \(\mc{C}_p^{\ms{a}}\) are called \textbf{discardable}, while morphisms
  in \(\mc{C}_1^{\ms{a}}, \mc{C}_1^{\ms{r}}, \mc{C}_1^{\{\ms{a},
  \ms{r}\}}\) specifically are called \textbf{affine}, \textbf{relevant}, and
  \textbf{intuitionistic} or \textbf{pure} respectively.
\end{definition}
We can obtain an example of a substructural effectful category quite easily by,
considering the power-set monad above, applying the state transformer to yield
the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to
\mc{P}(A \times S)\). Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
\(\upg{R}{} = \lambda a, s. R(a) \times \{s\}\), we can verify that this
satisfies the axioms of a substructural effectful category. Note that here the
\textit{central} morphisms are precisely those which do not read from or write
to the state, and therefore which commute with all other morphisms.

@@

We can now interpret a term \(\hasty{\Gamma}{p}{a}{A}{q}\) as a
\textit{morphism} taking \(\dnt{\Gamma}\) into \(\dnt{A}\) with
\textit{centrality} \(p\) and \textit{quantity} \(q\) (i.e., lying in
\(\mc{C}_p^q\)). This means that, as expected, a central term becomes a central
morphism in \(\mc{C}_1\), an affine term becomes an affine morphism, and a
relevant term becomes a relevant morphism, and so, in particular, a pure term
becomes a pure morphsim. Our rules for doing so are given in Figure
\ref{fig:ssa-term-semantics}; we will abuse syntax slightly and write
\(\dnt{\text{"judgement"}}\) for \(\dnt{D}\), where \(D\) is a derivation of
"judgement"; in general, it can be proven that the semantics of any two such
derivations are equal.
% \begin{lemma}[Coherence] \
%   \begin{itemize}
%     \item Given any two derivations \(D_1: \hasty{\Gamma}{p}{a}{A}{q}\), \(D_2: \hasty{\Gamma}{p}{a}{A}{r}\), \(\dnt{D_1} = \dnt{D_2}\)
%     \item Given any two derivations \(D_1, D_2: \haslb{\Gamma}{t}{\ms{L}}\), \(\dnt{D_1} = \dnt{D_2}\)
%     \item Given any two derivations \(D_1, D_2: \csplits{\Gamma}{\Delta}{\Xi}\), \(\dnt{D_1} = \dnt{D_2}\)
%     \item Given any two derivations \(D_1, D_2: \cwk{\Gamma}{\Delta}\), \(\dnt{D_1} = \dnt{D_2}\)
%     \item Given any two derivations \(D_1, D_2: \lwk{\ms{L}}{\ms{K}}\), \(\dnt{D_1} = \dnt{D_2}\)
%   \end{itemize}
% \end{lemma}
Our treatment of \rle{var} is quite standard: the term \(x\) types in \(\Gamma\)
if all other variables in \(\Gamma\) and can be discarded, in which case the
semantics is simply doing so, i.e. \(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\),
lifting to \(\mc{C}_0\) as necessary. Similarly, to type a function application
\rle{var}, we first take \(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is
guarnateed to be central, lift it to the appropriate category \(\mc{C}_p\), and
then simply postcompose it with the denotation of \(f\). \rle{pair} is handled
by composing the context split \(\csplits{\Gamma}{\Delta}{\Xi}\) taking the
tensor product of the components \(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\),
\(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is well-defined since both components
are central and \(\mc{C}_1\) is monoidal. We then lift the result to the
appropriate \(\mc{C}_p\). The semantics of \rle{unit} is simply dropping the
entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\), while \rle{true}, and \rle{false}
are each handled by first dropping the context and then postcomposing with the
appropriate constant; in all cases, this is followed by lifting to \(\mc{C}_p\).
Finally, \rle{let} and \rle{let2} are handled by first splitting the context,
then passing the right hand component to the semantics of the bound term
\(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or \(\dnt{\hasty{\Xi}{1}{a}{A \otimes
B}{q}}\), and then, lifting to \(\mc{C}_p\) as necessary, after reassociating
with \(\alpha\), passing everything to the inner term \(\dnt{\hasty{\Delta,
\thyp{x}{A}{q}}{p}{e}{B}{q}}\) or \(\dnt{\hasty{\Delta, \thyp{x}{A}{q},
\thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that this inner term is not necessarily
central, unlike the bound term!

>>>>

One issue with the above definition is that, just like for regular premonoidal
categories, it is difficult to distinguish between those morphisms that are
"truly" central versus those which are just central "by accident", as we only
want to require any notion of functor we develop subsequently to preserve the
former. We can solve this the same way, by simply extending the notion of
effectful category appropriately as follows:
\begin{definition}[Susbtructural Effectful Category]
  An effectful category \(\mc{C}_0 \to \mc{C}_1\) is said to be
  \textbf{substructural} if \(\mc{C}_0, \mc{C}_1\) are substructural and
  \(\upg{\cdot}{}: \mc{C}_1^\varnothing \to \mc{C}_0^\varnothing\) preserves all
  symmetries, associators, unitors, drops, and splits and sends \(\mc{C}_1^q\)
  to \(\mc{C}_0^q\). In particular, restricting \(\upg{\cdot}{}\) appropriately
  makes all \(\mc{C}_1^q \to \mc{C}_0^q\) into effectful categories. Morphisms
  in \(\mc{C}_p^{\ms{r}}\) are called \textbf{copyable}, morphisms in
  \(\mc{C}_p^{\ms{a}}\) are called \textbf{discardable}, while morphisms in
  \(\mc{C}_1^{\ms{a}}, \mc{C}_1^{\ms{r}}, \mc{C}_1^{\{\ms{a}, \ms{r}\}}\)
  specifically are called \textbf{affine}, \textbf{relevant}, and
  \textbf{intuitionistic} or \textbf{pure} respectively.
\end{definition}
We can obtain an example of a substructural effectful category quite easily by,
considering the power-set monad above, applying the state transformer to yield
the Kliesli category with morphisms of the form \(\mc{C}_0(A, B) = A \to S \to
\mc{P}(A \times S)\). Letting \(\mc{C}_1^q = \ms{Rel}^q\) and defining
\(\upg{R}{} = \lambda a, s. R(a) \times \{s\}\), we can verify that this
satisfies the axioms of a substructural effectful category. Note that here the
\textit{central} morphisms are precisely those which do not read from or write
to the state, and therefore which commute with all other morphisms.

@@

We can now interpret a term \(\hasty{\Gamma}{p}{a}{A}{q}\) as a
\textit{morphism} taking \(\dnt{\Gamma}\) into \(\dnt{A}\) with
\textit{centrality} \(p\) and \textit{quantity} \(q\) (i.e., lying in
\(\mc{C}_p^q\)). This means that, as expected, a central term becomes a central
morphism in \(\mc{C}_1\), an affine term becomes an affine morphism, and a
relevant term becomes a relevant morphism, and so, in particular, a pure term
becomes a pure morphsim. Our rules for doing so are given in Figure
\ref{fig:ssa-term-semantics}; we will abuse syntax slightly and write
\(\dnt{\text{"judgement"}}\) for \(\dnt{D}\), where \(D\) is a derivation of
"judgement"; in general, it can be proven that the semantics of any two such
derivations are equal. We also define the syntax sugar \(\upg{f}{p}\), where
\(\upg{f}{0}\) is \(\upg{f}{}\) for morphisms in \(\mc{C}_1\) and simply \(f\)
for morphisms in \(\mc{C}_0\), and \(\upg{f}{1}\) is the identity functor on
\(\mc{C}_1\).

Our treatment of \rle{var} is quite standard: the term \(x\) types in \(\Gamma\)
if all other variables in \(\Gamma\) and can be discarded, in which case the
semantics is simply doing so, i.e. \(\dnt{\cwk{\Gamma}{\thyp{x}{A}{q}}}\),
lifting to \(\mc{C}_0\) as necessary. Similarly, to type a function application
\rle{var}, we first take \(\dnt{\hasty{\Gamma}{1}{a}{A}{q}}\), which is
guarnateed to be central, lift it to the appropriate category \(\mc{C}_p\), and
then simply postcompose it with the denotation of \(f\). \rle{pair} is handled
by composing the context split \(\csplits{\Gamma}{\Delta}{\Xi}\) taking the
tensor product of the components \(\dnt{\hasty{\Delta}{1}{a}{A}{q}}\),
\(\dnt{\hasty{\Xi}{1}{b}{B}{q}}\); this is well-defined since both components
are central and \(\mc{C}_1\) is monoidal. We then lift the result to the
appropriate \(\mc{C}_p\). The semantics of \rle{unit} is simply dropping the
entire context; \(\dnt{\cwk{\Gamma}{\cdot}}\), while \rle{true}, and \rle{false}
are each handled by first dropping the context and then postcomposing with the
appropriate constant; in all cases, this is followed by lifting to \(\mc{C}_p\).
Finally, \rle{let} and \rle{let2} are handled by first splitting the context,
then passing the right hand component to the semantics of the bound term
\(\dnt{\hasty{\Xi}{1}{a}{A}{q}}\) or \(\dnt{\hasty{\Xi}{1}{a}{A \otimes
B}{q}}\), and then, lifting to \(\mc{C}_p\) as necessary, after reassociating
with \(\alpha\), passing everything to the inner term \(\dnt{\hasty{\Delta,
\thyp{x}{A}{q}}{p}{e}{B}{q}}\) or \(\dnt{\hasty{\Delta, \thyp{x}{A}{q},
\thyp{y}{B}{q}}{p}{e}{C}{q}}\). Note that this inner term is not necessarily
central, unlike the bound term!
